---
title: "Exploring Indian Weather Patterns: Insights from Data"
output: html_document
author: Jeffrey Asubonteng
date: "2023-11-27"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Introduction

This report focuses on exploring Indian weather dataset. This dataset can be found
on Kaggle website. It is about 7MB, consisting of 28448 observations and 42 variables, some of which include temperature, humidity, region, etc. The goal of this project is to apply what I have
learnt in my datascience class to explore the dataset. In this report, I apply Principal
Component Analysis, t-SNE, clustering, supervised learning, and data visualization on the dataset. With these statistical method mentioned, I apply them on some selected variables.

```{r}

library(readr)
IndianWeatherRepository <- read_csv("./source_data/IndianWeatherRepository.csv")

IndianWeather <- IndianWeatherRepository
```

### Data Overview and Analysis
* Temperature trends over time

Temperature variable in the dataset are in both celcius and farenheit so I chose the temperature-celcius variable against the last time the dataset was updated. From the figure,
the teperature was well distributed. 


```{r}
library(ggplot2)

ggplot(IndianWeather, aes(x = last_updated, y = temperature_celsius)) +
  geom_line() +
  labs(title = "Temperature Trends over Time", x = "Date", y = "Temperature (°C)")
```

* Distribution of weather conditions

Among the observations, there are different weather conditions in the dataset so I considered to visualize the counts. From the count, it is observed that India has more clear weather condition compared to the others.
```{r}
# Weather Conditions
library(ggplot2)
ggplot(IndianWeather, aes(x = condition_text)) +
  geom_bar() +
  coord_flip() +
  labs(title = "Weather Conditions", x = "Count", y = "Condition")
```


