---
title: "Exploring Indian Weather Patterns: Insights from Data"
output: html_document
author: Jeffrey Asubonteng
date: "2023-11-27"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Introduction

This report focuses on exploring Indian weather dataset. This dataset can be found
on Kaggle website. It is about 7MB, consisting of 28448 observations and 42 variables, some of which include temperature, humidity, region, etc. The goal of this project is to apply what I have
learnt in my datascience class to explore the dataset. In this report, I apply Principal
Component Analysis, t-SNE, clustering, supervised learning, and data visualization on the dataset. With these statistical method mentioned, I apply them on some selected variables.

```{r}

library(readr)
IndianWeatherRepository <- read_csv("./source_data/IndianWeatherRepository.csv")

IndianWeather <- IndianWeatherRepository
```

### Data Overview and Analysis

* Temperature trends over time

Temperature variable in the dataset are in both celsius and farenheit so I chose the temperature-celsius variable against the last time the dataset was updated. From the figure,
the teperature was well distributed. 


```{r}
library(ggplot2)

ggplot(IndianWeather, aes(x = last_updated, y = temperature_celsius)) +
  geom_line() +
  labs(title = "Temperature Trends over Time", x = "Date", y = "Temperature (°C)")
```

* Distribution of weather conditions

Among the observations, there are different weather conditions in the dataset so I considered to visualize the counts. From the count, it is observed that India has more clear weather condition compared to the others.

```{r}
# Weather Conditions
library(ggplot2)
ggplot(IndianWeather, aes(x = condition_text)) +
  geom_bar() +
  coord_flip() +
  labs(title = "Weather Conditions", x = "Count", y = "Condition")
```

* Geographic temperature variation

I delve into the analysis of temperature variations across locations in India. 
Utilizing statistical visualization I aim to elucidate the spatial distribution of temperatures
throughout the country. The code below constructs a geographical scatter plot depicting temperature fluctuations across longitude and latitude coordinates in India. The visualization
represents temperature values as a color gradient, allowing for an intuitive 
comprehension of temperature disparities across various geographical regions.

```{r}
library(ggplot2)
ggplot(data = IndianWeather, aes(x = longitude, y = latitude, color = temperature_celsius)) +
  geom_point() +
  scale_color_gradient(low = "blue", high = "red", name = "Temperature (Celsius)") +
  labs(title = "Temperature Variation by Location in India")

```

* Other multivariate analysis using PCA and clustering techniques

Principal component analysis reduces data dimensionality, revealing the variance captured 
by PC1 and PC2. The code below visualizes PC1 vs. PC2 incorporating K-means clustering
to identify data groupings.

```{r}
selected_variables <- IndianWeather[, c("temperature_celsius", "humidity", "wind_mph")]
scaled_data <- scale(selected_variables) 

pca_result <- princomp(scaled_data)
# View the loadings for the first two principal components
loadings(pca_result)[, 1:2]

# Extract PC scores
pc_scores <- pca_result$scores


# Clustering
num_clusters <- 3
kmeans_result <- kmeans(scaled_data, centers = num_clusters)

cluster_assignments <- kmeans_result$cluster
table(cluster_assignments)


# Assuming PC1 and PC2 are the first two principal components
plot(pc_scores[, 1], pc_scores[, 2], 
     col = cluster_assignments, pch = 16,
     xlab = "Principal Component 1", ylab = "Principal Component 2",
     main = "PC1 vs. PC2")
legend("topright", legend = unique(cluster_assignments), col = 1:length(unique(cluster_assignments)), pch = 16)
```

* Correlation Analysis

To better understand the relationships between wind speed, humidity, and sunrise, correlation coefficients were computed. The correlation table reveals the strength and direction of 
associations between these variables.

```{r}
relevant_data <- IndianWeather[, c("wind_mph", "humidity", "sunrise")]

# Convert time of sunrise to numeric format (e.g., minutes since midnight)
relevant_data$sunrise_numeric <- as.numeric(relevant_data$sunrise)
# Selecting relevant columns for correlation
relevant_columns <- relevant_data[c("wind_mph", "humidity", "sunrise_numeric")]

# Calculating correlation matrix
correlation_matrix <- cor(relevant_columns)

correlation_matrix
```
* Heatmap Visualization.

A heatmap representation of these correlations visually displays the relationships
among the variables. The graphical illustration aids in the easy identification of 
varifying degrees of association between wind speed, humidity, and sunrise.

```{r}
library(ggplot2)
library(reshape2)
correlation_matrix <- data.frame(
  wind_mph = c(1.00000000, -0.01813135, 0.24702198),
  humidity = c(-0.01813135, 1.00000000, -0.42345699),
  sunrise_numeric = c(0.247022, -0.423457, 1.000000)
)

# Redefine the melted matrix with variable names specified
melted_matrix <- melt(correlation_matrix)

ggplot(melted_matrix, aes(x = variable, y = variable, fill = value)) +
  geom_tile() +
  scale_fill_gradient(low = "blue", high = "red") +
  labs(title = "Correlation Heatmap") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

* PCA and t-SNE analysis

Principal Component Analysis and t-distributed Stochastic Neighbor Embedding were conducted 
on the selected variables. The results, visually represent, reveal the underlying 
structures and relationships among these variables in reduced dimensions. These visualizations
assist in comprehending complex interrelationships among the variables, offering
insights into their mutual associations and potential patterns.

```{r}
library(dplyr)
IndianWeather_subset <- IndianWeather %>%
  mutate(sunrise_seconds = as.numeric(IndianWeather$sunrise)) %>%
  select(wind_mph, humidity, sunrise_seconds)

# Assuming 'IndianWeather_subset' contains your data
pca_result1 <- prcomp(IndianWeather_subset, scale. = TRUE)
summary(pca_result1)

library(ggplot2)
# Convert PCA result to a data frame
pca_df <- as.data.frame(pca_result1$x[, 1:2])  # Considering the first two principal components
ggplot(pca_df, aes(x = PC1, y = PC2, color = IndianWeather_subset$wind_mph)) +
  geom_point() +
  labs(title = "PCA Plot with Color") +
  theme_minimal()

# TSNE

library(Rtsne)

IndianWeather_unique <- unique(IndianWeather_subset)

tsne_result <- Rtsne(IndianWeather_unique, dims = 2, perplexity = 30, verbose = TRUE)
tsne_df <- as.data.frame(tsne_result$Y)

ggplot(tsne_df, aes(x = V1, y = V2)) +
  geom_point() +
  labs(title = "t-SNE Plot") +
  theme_minimal()
```

* Supervised Learning

I delved into supervised learning, employing random forest model to predict tempeature 
from the dataset. After splitting the data into training and testing set, I trained the 
model using root mean squared error. Below is the code and plot generated from the model.


```{r}
X <- IndianWeather_subset
y <- IndianWeather$temperature_celsius  # Target variable

#Train-Test split

library(caret)
set.seed(123)  # For reproducibility
trainIndex <- createDataPartition(y, p = 0.7, list = FALSE)
X_train <- X[trainIndex, ]
X_test <- X[-trainIndex, ]
y_train <- y[trainIndex]
y_test <- y[-trainIndex]

# Model selection and training
library(randomForest)
model <- randomForest(y_train ~ ., data = X_train)

# Prediction
predictions <- predict(model, newdata = X_test)


library(Metrics)
rmse_value <- rmse(predictions, y_test)

# Visualization

library(ggplot2)

# Create a data frame with actual and predicted values
data <- data.frame(Actual = y_test, Predicted = predictions)

# Plotting actual vs predicted values
ggplot(data, aes(x = Actual, y = Predicted)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +  # Adding a diagonal line for reference
  labs(title = "Actual vs Predicted of temperature", x = "Actual Values", y = "Predicted Values") +
  theme_minimal()
```

### Conclusion

In this project, I observed that temperature from the last time the dataset was updated 
did not varied that much. Temperature was normally between -9 to 38 degrees celsius. 
There were various weather condition in India. Per count, the weather condition is mostly clear, followed by mist and partly cloudy weather. The principal component analysis that I performed on Temperature, Humidity and wind speed, the clusters were difficult to identified except the color improved the visualization. Again, PCA on  wind_mph, humidity, sunrise_seconds
did not generate anything interesting. But the t-SNE visualization on wind_mph, humidity, sunrise_seconds looks aesthetically pleasing. Lastly, the supervised learning performed on temperature generated a better model. This is observed from the actual vs predicted temperature plot as the temperature data point are clustered as long the line.

This dataset contains a lot of variables which are meaningful but I chose to focused on few variables like temperature, humidity, sunrise and others. I have not found anything interesting so far except for the t-SNE analysis which I don't understand how it generated the figure, but as time goes on in my analysis I hope to achieve something educative.
